---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}

#unzip the file
file <- unzip("activity.zip")
#read the file
df1 <- read.csv(file, header = TRUE, sep = ",", na.strings = "NA", nrows = 17568)

#assingn  to another data frame so we can use for later analysis 
df2 <- df1

#geta clean dataframe without NAs
df3 <- na.omit(df1)


```

## What is mean total number of steps taken per day?

```{r 1.Histogram }
library(sqldf)
Tot <- sqldf("select sum(steps) from df3 group by date ")
colnames(Tot) <- "steps"
```
#### 1. Total number of steps taken per day
```{r}
print(Tot)
```

#### 2. Histogram of total number of steps taken
```{r}
par(mfrow = c(1,1))

hist(Tot$steps , breaks = 25, ylim = c(0,10), col = 'brown', xlab = 'Sum of Steps', bg = 'white', main = 'Number of Steps per day')


```

```{r 2.Mean and Median}
MMPDay <- sqldf("select date as Date, AVG(steps) as Mean, ((max(steps) + min(steps)) / 2)  as Median from df3 group by date ")
```

#### 3.Mean and Median of the total number of steps taken.
```{r}
print(MMPDay)

```

## What is the average daily activity pattern?
```{r Average Daily Activity Pattern}

library(lattice)

AveStepsInt <- sqldf("select interval, AVG(steps) as Mean from df3 group by interval ")
```
#### 1. Time series plot of 5 min interval and average steps taken.
```{r}

xyplot(AveStepsInt$Mean ~ AveStepsInt$interval,
        data = AveStepsInt,
        type = "l", lty = c(1, 2, 2, 1),
        xlab = "Interval", ylab = "Average Steps"
       )

```

####2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r Interval at Maximum number of steps}
maxInt <- sqldf(" select interval from df3 where steps = (select MAX(steps) from df3)")

print(maxInt)

```

## Imputing missing values
### Strategy is to to fill with mean of interval for NA's while counting them and create a new dataset with missing values filled in (this covers 2 & 3 of the instruction for assignment)
```{r Impute}

TotalNas <- 0

for(i in 1:17568) {
        if (is.na(df1$steps[i])) {  
            m =df1$interval[i]
            df2$steps[i] <- as.integer(sqldf(paste("select Mean from AveStepsInt where interval = " ,m)))
            TotalNas <- TotalNas + 1
            
        }
}

colnames(df2) <- c("steps", "date", "interval")

```
#### 1. Total Number of NA values
```{r}
print(TotalNas)

```
### 4. Histogram after Imputing
```{r 2.Histogram }

Tot2 <- sqldf("select sum(steps) from df2 group by date ")
colnames(Tot2) <- "steps"

par(mfrow = c(1,1))

hist(Tot2$steps , breaks = 25,, ylim = c(0,10), col = 'brown', xlab = 'Sum of Steps', bg = 'white', main = 'Number of Steps per day')

```

### 4. Mean and Median after imputing
```{r Mean and Median}
MMPDay2 <- sqldf("select date as Date, AVG(steps) as Mean, ((max(steps) + min(steps)) / 2)  as Median from df2 group by date ")

MMPDay2$Median <- as.integer(MMPDay2$Median)

print(MMPDay2)

```
## Impact after imputing missing date

#### Surprisingly the graph, mean and median hasnt changed at all
#### well it make sense since we are filling in with mean value which shouldnt really affect the median/mean/graph. 



## Are there differences in activity patterns between weekdays and weekends?

```{r Differences in activity patters between weekdays and weekends}

#create a vector of weekdays
df2$day <- weekdays(as.Date(df2$date))
weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
#Use `%in%` and `weekdays` to create a logical vector
#convert to `factor` and specify the `levels/labels`
df2$wDay <-  factor((df2$day %in% weekdays1)+1L,
      levels=1:2, labels=c('weekday', 'weekend'))

```

## Making a time series panel plot for weekdays versus weekends

#### 1. Creating a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r time series panel plot}

library(lattice)
AveStepsInt2 <- sqldf("select interval,
                      wDay,
                      AVG(steps) as Mean 
                      from df2 
                      where wDay = 'weekday'
                      group by interval ")
AveStepsInt3 <- sqldf("select interval,
                      wDay,
                      AVG(steps) as Mean 
                      from df2 
                      where wDay = 'weekend'
                      group by interval ")


df4 <- rbind(AveStepsInt3, AveStepsInt2)
````

#### 2. Panel Plot 5 min interval and average number of steps of taken.

```{r Panel Plot}

xyplot(Mean ~ interval | wDay, 
        data = df4,
        layout=c(1,2),
        type = "l", lty = c(1, 2, 2, 1),
        xlab = "Interval", ylab = "Number of Steps"
       )


```

